---
title: High School and Beyond (HSB)
---

## Dataset Description 

A subset of data from the 1982 [High School and Beyond (HSB) survey](<https://nces.ed.gov/surveys/hsb/>){target="_blank"} used as examples for HLM software.

```{r}
#| label: dataset-help
#| code-line-numbers: false
#| output-location: default
#| echo: true
#| eval: false

?merTools::hsb
```

A data frame with 7,185 observations on the following 8 variables.


:::: {.columns}

::: {.column width="49%"}

- `schid`: a numeric vector, 160 unique values
- `mathach`: a numeric vector for the performance on a standardized math assessment
- `female`: a numeric vector coded 0 for male and 1 for female
- `ses`: a numeric measure of student socio-economic status

:::

::: {.column width="2%"}

:::

::: {.column width="49%"}


- `minority`: a numeric vector coded 0 for white and 1 for non-white students
- `schtype`: a numeric vector coded 0 for public and 1 for private schools
- `meanses`: anumeric, the average SES for each school in the data set
- `size`: a numeric for the number of students in the school

:::

::::

## Import HSB dataset 

```{r}
#| label: dataset
#| echo: true
#| output-location: fragment
#| code-line-numbers: "1|2"
dat <- merTools::hsb
head(dat, 10)

```

