---
title: Simulated Data based on 3-dimensional CFA Multiple Group Model 
---

## Dataset Description {#cfa-pre-dat-1}

A simulated dataset used for examples in Structural Equation Modeling. The data generating model is a 3-dimensional CFA multiple group model with correlated error variables of the item indicators (sometimes called Latent state model). 


```{r}
#| label: sim-ls-data-0
#| echo: true
#| code-copy: true
#| code-annotations: false
#| code-line-numbers: 2|3-5|6-8|9-14|15-16|17-19|20-23|
#| output-location: column
# <0>
popLsModel <- ' # <1>
eta1 =~ .6*Y11 + .6*Y21 + .6*Y31 # <2>
eta2 =~ .6*Y12 + .6*Y22 + .6*Y32 # <2>
eta3 =~ .6*Y13 + .6*Y23 + .6*Y33 # <2>
Y11 + Y12 + Y13 ~ i1*1 # <3>
Y21 + Y22 + Y23 ~ i2*1 # <3>
Y31 + Y32 + Y33 ~ i3*1 # <3>
Y11 ~~ .15*Y12 + .1*Y13 # <4>
Y12 ~~ .15*Y13 # <4>
Y21 ~~ .15*Y22 + .1*Y23 # <4>
Y22 ~~ .15*Y23 # <4>
Y31 ~~ .15*Y32 + .1*Y33 # <4>
Y32 ~~ .15*Y33 # <4>
eta1 ~~ .5*eta2 + .35*eta3 # <5>
eta2 ~~ .5*eta3 # <5>
eta1 ~ c(0,0)*1 # <6>
eta2 ~ c(0.25, 0.25)*1 # <6>
eta3 ~ c(0.5, 0.75)*1 # <6>
eta1 ~~ c(1,1)*eta1 # <7>
eta2 ~~ c(1,1)*eta2 # <7>
eta3 ~~ c(1,1.5)*eta3 # <7>
'
```


1. Define the population model within a literal string (i.e., surrounded by `' '`)
2. Specifiy measurement models (`=~`) and factor loadings (`*`)
3. Specify intercepts of item indicators (`~ 1`)
4. Specify error covariances (`~~`)
5. Specify covariances between latent variables (`~~`)
6. Specify means of latent variables (`~ 1`)
7. Specify variances of latent variables (`~ 1`)



## Simulating the Data

<br>

```{r}
#| label: sim-ls-data-1
#| echo: true
#| code-annotations: below
#| code-line-numbers: 2|3|4|5|6|
#| output-location: column

# <0>
wideLSdat <- lavaan::simulateData( # <1>
    model = popLsModel, # <2>
    meanstructure = TRUE, # <3>
    sample.nobs = c(500, 500), # <4>
    seed = 987) # <5>

wideLSdat$group <- wideLSdat$group-1
```

1. To generate the dataset with the name `wideLSdat`, we use the `simulateData` function from the `lavaan` package.
2. In the model argument, we state the generated string (`popLsModel`) from above (the previous [slide](#cfa-pre-dat-1)).
3. (Optional) Setting the `meanstructure` argument to `TRUE`.
4. The `sample.nobs` argument controls the number of observation (per group). Here 2 groups with $n$ = 500.
5. Defining `seed` for reproducible results.


```{r}
#| label: save-wideLSdat
#| eval: false
saveRDS(wideLSdat, "data/wideLSdat.RDS")
```