---
title: "Functions and Loops"
author: "Sven Rieger"
published-title: "Last update on"
date: today
toc: true
---



## Writing functions

Writing functions is the best way to enhance your data processing skills. Functions allow you to automate tasks that are needed to be repeated more than 2-times. This sections gives a brief introduction in writing your own functions. There are other sources that cover this topic in much more detail (see [R for Data Science](https://r4ds.had.co.nz/functions.html?q=functions#functions) by [Hadley Wickham](https://hadley.nz/) & [Garrett Grolemund](https://www.rstudio.com/authors/garrett-grolemund/)). 

Example: We want to calculate the mean, standard deviation and range of two variables. 

The copy/paste way.

```{r}
#| label: own-function-1
#| results: hide
#| code-line-numbers: true

set.seed(999)

x <- rnorm(50)
y <- rnorm(50)

mean(x)
sd(x)
range(x)

mean(y)
sd(y)
range(y)
```

Lets put it a function.

1. Pick a name for the function (here: ```myFirstFunction```) 
2. Define the inputs and arguments of the function. These are written within the regular brackets ```(...)``` (here the input is: variable) 
3. Write the code in the body of the function which is between the curly brackets ```{ }```.
    - Calculate the mean, the standard deviation and the range
    - Store it in a named list (for what a named list is useful see [below](#list-chr))
    - ```return``` the named list

```{r}
#| label: own-function-2

myFirstFunction <- function ( variable ) {
  
  fMEAN <- mean(variable)
  fSD <- sd(variable)
  fRANGE <- range(variable)
  
  fOut <- c("Mean" = fMEAN,
            "SD" = fSD,
            "Min" = fRANGE[1],
            "Max" = fRANGE[2])
  
  return(fOut)
  
}


myFirstFunction(variable = x)
```


## Loops {#loops}


some text about loops

To execute the function on both variables...

::: {.panel-tabset}

### Approach 1: apply family

```{r}
#| label: loop-1


sapply(list(x, y),
       function(myVar) myFirstFunction(variable = myVar),
       simplify = F)

```

### Approach 2: for loop

```{r}
#| label: loop-2

for (myvar in list(x,y)) {
  print(
    myFirstFunction(variable = myvar)
  )
}
```

:::

